Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-07 demystify.tex
updates

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing your admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your 5--6 year PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost}
          \item Don't buy computer equipment and electronics for research, ask your prof. if they can buy them for you (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

In my personal opinion, the short answer is \emph{yes}, it \emph{does} help, especially if you can articulate your experiences and how you can contribute to diversity and inclusion in CS. Most universities have a strong commitment to diversity and inclusion and want to  
\emph{increase the number of URM students} in their programs (e.g., GMU is consistently ranked as one of the most diverse universities in the US). Moreover, many universities and industrial organizations also have specific scholarships and fellowships for URM students (so you should look for specific scholarships for URM students).

@@

In addition to being aware of the importance of diversity and inclusion in CS and want to contribute to it, profs. have incentives to work with URM students. For example, professors often seek grants from government sponsors such as NSF, which strongly encourage broadening participation in STEM (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}). If in doubt, seek advice from your close professors and collaborators about this.

@@

\begin{table}
  \centering
  \small
  \caption{GRA cost breakdown. F \& A is Facilities \& Administrative Cost Base and
    MTDC is Modified Total Direct Cost. These are things that the university can charge overhead to.}\label{tab:cost}
  \begin{tabular}{rcl}
    \textbf{Budget} & \textbf{Cost \$} & \textbf{Notes} \\
    \midrule
    GRA (9-month) & 27K & \\
    GRA (summer)  &9K	& 3-month, 20hrs/week\\
    \textbf{Total Salary} &\red{36K}	&\\
    \midrule
    Health Insurance	&3K	& full year\\
    Tuition (In-State) &	15K	& (\$680/ Credit + \$150/Student Fee/ Credit)* 9 credits = \\
                    &&\$7470 (\$6120 + \$1350) per semester\\
    \textbf{Total Tuition \& Insurance}	&\red{18K}	&Full year tuition + insurance\\
    \midrule
    Conference Registration	& 500 & \\
    International Travel &	1800& \\
    Domestic Travel	& 700	& \\
    \textbf{Total Travel}&	\red{3K}	\\
    \midrule
    Total Direct Cost	& \red{57K}	&Salary + Travel + Health + Tuition \\
    F \& A (MTDC)	& 21K	& Direct Cost - GRA Salary\\
    Total Indirect Cost	& \red{12K}	&58.9\% of MTDC\\
    \textbf{Total (Direct + Indirect)} &	\red{69K}	& Budget for a GRA\\
    \bottomrule
  \end{tabular}
\end{table}

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, in this table I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

At the end, the total budget comes out to be \$69K/year to support a PhD student. \emph{The summary is that over your 5-6 year of your PhD, you cost about \$350K--400K, and while you're paid X, your adviser probably pays 2X for you}.

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} Many students, including myself, get very excited about getting to buy new computer equipment and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you a computer and other equipments (e.g., keyboards, headphones, tablets, books, etc).

\section{Can I negotiate my PhD offer?}\label{sec:negotiate}
For stipend, unlikely, as TA is set by the university and while RA is set by the adviser, it is often based on the department's standard and it would be unfair to other students if you get more. However, your stipend is often increased every year (by a small amount and automatically) and will be more after you pass your qualifying exam.

For start date or TA assignment (e.g., TA a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, things that are standard and set by the university or department are unlikely to be negotiable.  However, you can ask for things such as computer equipments (\autoref{sec:buying-equipment}), books, (and sometimes even furniture) and conference travel budget.

@@

Most of what is written in this document is applicable to both domestic\footnote{Domestic simply means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences that domestic students should be aware of.  This section lists some of these differences.

@@

\paragraph{Standing out} There are \emph{very few} domestic applications compared to international ones (i.e., domestic students are minority in the CS PhD application pool), and many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

>>>>

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in the underepresentative group in CS can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipments and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google) (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

In my personal opinion, the short answer is \emph{yes}, it \emph{does} help, especially if you can articulate your experiences and how you can contribute to diversity and inclusion in CS. Most universities have a strong commitment to diversity and inclusion and want to  
\emph{increase the number of URM students} in their programs (e.g., GMU often advertises about being one of the most diverse universities in the US). Moreover, many universities and industrial organizations also have specific scholarships and fellowships for URM students (so you should look for specific scholarships for URM students).

@@

In addition to being aware of the importance of diversity and inclusion and want to contribute to it, profs. have incentives to work with URM students. For example, professors often seek grants from government sponsors such as NSF, which strongly encourage broadening participation in STEM (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}). Prof. can also apply for grants from organizations and private donors, which have specific programs for URM students.  Thus, having URM students in their group can help them get these grants.

In the end, this is your choice and you do what make you feel comfortable. However, just be aware that URM is often a (huge) plus, not a minus, in CS in the US. If in doubt, seek advice from your close professors and collaborators about this.

@@

\begin{table}
  \centering
  \small
  \caption{GRA cost breakdown. F \& A is Facilities \& Administrative Cost Base and
    MTDC is Modified Total Direct Cost. These are things that the university can charge overhead to.}\label{tab:cost}
  \begin{tabular}{rcl}
    \textbf{Budget} & \textbf{Cost \$} & \textbf{Notes} \\
    \midrule
    GRA (9-month) & 29K & \\
    GRA (summer)  &10	& 3-month, 20hrs/week\\
    \textbf{Total Salary} &\red{39K}	&\\
    \midrule
    Health Insurance	&3K	& full year\\
    Tuition (In-State) &	15K	& (\$680/ Credit + \$150/Student Fee/ Credit)* 9 credits = \\
                    &&\$7470 (\$6120 + \$1350) per semester\\
    \textbf{Total Tuition \& Insurance}	&\red{18K}	&Full year tuition + insurance\\
    \midrule
    Conference Registration	& 500 & \\
    International Travel &	1800& \\
    Domestic Travel	& 700	& \\
    \textbf{Total Travel}&	\red{3K}	\\
    \midrule
    Total Direct Cost	& \red{60}	&Salary + Travel + Health + Tuition \\
    F \& A (MTDC)	& 21K	& Direct Cost - GRA Salary\\
    Total Indirect Cost	& \red{12K}	&58.9\% of MTDC\\
    \textbf{Total (Direct + Indirect)} &	\red{72K}	& Budget for a GRA\\
    \bottomrule
  \end{tabular}
\end{table}

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$29000 and a 3-month summer of \$10000 (about a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

At the end, the total budget comes out to be \$72K/year to support a PhD student. \emph{The summary is that over your 5--6 years of your PhD, you cost about \$400K, and while your stipend is X, your adviser probably pays 2X for you}.

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} Many students get very excited about getting to buy new computer equipment and electronics when they start their PhD. However, you should check with your adviser first.  Many profs. have funding to buy you a computer and other equipments (e.g., keyboards, headphones, tablets, books, etc).

\section{Can I negotiate my PhD offer?}\label{sec:negotiate}
For stipend, unlikely, as TA is set by the university and while RA is set by the adviser, it is often based on the department's standard, and it also would be unfair to other students if you get more. However, your stipend is often increased every year (by a small amount and automatically) and will likely increase after you pass your qualifying exam.

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, things that are standard and set by the university or department are unlikely to be negotiable.  However, you can ask for things such as computer equipments (\autoref{sec:buying-equipment}), books, (and sometimes even furniture) and conference travel budget.

@@

Most of what is written in this document is applicable to both domestic\footnote{Domestic simply means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of.  This section lists some of these.

@@

\paragraph{Standing out} There are \emph{very few} domestic applications compared to international ones, i.e., domestic students are minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

\paragraph{Open House} (\autoref{sec:accepted}) It is easier for you to attend open house events in person.  This can help you make a better decision on which school to attend.
